AB REP Surplus from Hastings-Simon et al (2022)

### 2025-02-24 paper/paper_draft.tex
paper edits

%%%% AUTHOR INFORMATION %%%%
% AUTHOR INFO
\author{Andrew Leach\\
 University of Alberta
 \and
Blake Shaffer\\University of Calgary}
\begin{document}
\maketitle

@@

We evaluate the response of wholesale power market participants in Alberta to a series of changes in greenhouse gas (GHG) emissions policies.  Between 2015 and 2019, legislative and regulatory changes have affected carbon price levels, the rates of output-based allocations of emissions credits, and the coverage of carbon pricing across facilities of different sizes, creating variation in average and marginal carbon prices both across and within facility types and across and within portfolios held by major players in Alberta's wholesale power market. We exploit this variance in treatment to provide unique evidence of carbon price pass-through in power markets. We show that even after adjusting for costs imposed by carbon policies at the facility level, statistically significant changes in portfolio-level offer curves were induced leading to a shift in the merit order toward gas and away from coal which exceeded that which would have been predicted by carbon prices alone. These changes have accelerated the phase-out of coal-fired power in the province. We also show, in contrast to much of the recent literature, that net marginal carbon price costs were not fully pass-through to offers. In fact, our results suggest that less than half of the changes in carbon policy costs was reflected in shifts in the merit order of electricity price offers in Alberta.

The traditional view of competitive firm response to carbon pricing would be an upward shift in each firm's supply function of an amount equal to the net incremental costs imposed by the policy. For some facilities, there may be dynamic gains to participating in the market in a given hour (for example, if shutdown implies multi-hour restarts, as can be the case for coal generators), in which case the supply curve may lie below zero up to some positive quantity supplied in each hour. For these facilities, especially if they are inframarginal, we might not expect to see a carbon price change their bidding behaviour up to some threshold quantity. However, above such a threshold, we should expect to see carbon price pass-through into power offers. However, our results in the Alberta context suggest less than perfect pass-through, even for units likely to be on the margin in the market in a given hour.

The degree to which carbon prices are passed-through to wholesale electricity markets has been extensively investigated. The studies of \cite{fabra_reguant} and \cite{hintermann16_germany} are most-closely related to our analysis, and both find that carbon costs imposed in the European Union emissions trading system are almost completely passed-through to market prices. In the case of \citeauthor{fabra_reguant}, they find more-than-complete (110\%) pass-through of carbon costs in peak hours, and less-than-complete (60\%) pass-through in off-peak hours. \citeauthor{hintermann16_germany} finds pass-through rates of 81-111\% depending on the hour of the day, and also finds that costs are fully passed-through on average. Our paper is unique in that we can measure facility-specific carbon policy costs (prices and the value of output-based allocations) precisely for long periods of time, which allows us to separately identify changes due to carbon pricing from changes due to other coincident market or climate factors. Furthermore, the changes in Alberta's climate change policies were such that the average and net marginal costs of carbon pricing policies changed across facilities for the same time periods. For example, in 2018, the marginal carbon price did not change from 2017, but there were substantial changes in output-based allocations of emissions credits such that marginal net carbon costs went up for some facilities and down for others. We use this variation to identify pass-through more precisely than is possible with daily market fluctuations in a cap-and-trade market. Our results estimate much lower pass-through of carbon prices to electricity offers and thus to market prices than either \cite{fabra_reguant} or \cite{hintermann16_germany}.

Average carbon policy costs varied over time and between fuel types in Alberta through the various changes to emissions policies since 2015. This links our work to \cite{cullen_mansur} which uses variations in fuel costs (coal vs. natural gas) to identify impacts on emissions. They find substantial impacts from small changes in fuel costs which make natural-gas-fired plants more competitive relative to coal plants. The policy changes we study in Alberta, both the changes in output-based allocations and the carbon price increases, have had a similar effect on the coal-to-gas cost ratio as a reduction in natural gas prices. Their results also inform our results in the sense that the same gas price reductions seen in their study occurred in Alberta during our sample period, and we adjust for these impacts in our study design. We show that the degree to which carbon prices were passed through was most consistent with the theory for more flexible, combined-cycle natural gas facilities, while coal plants passed through a much lower share of increased emissions costs.

The Alberta electricity market is highly concentrated, which links our paper to the literature following from the seminal paper of \cite{buchanan1969external} on externalities and related policies in the presence of market power, as well as to myriad studies of market power in electricity, e.g. \cite{borenstein_bushnell_wolak_2002}. For example, \cite{fowlie_reguant_ryan_2016} study the interaction of market structure in the cement industry and greenhouse gas mitigation policies. \cite{miller_cement}, also looking at the cement industry, shows that fuel cost changes were more-than-fully reflected in downstream pricing. Similarly, as in \cite{holland_emissions_2012}, incomplete regulation or regulations which provide competing incentives will distort price-pass through. The latter also applies in our work since there are changes in market structure coincident with the policy changes we study for which we must account.

Another unique attribute of our study is that the policy variation over time and across facilities in Alberta also allows us to directly identify the pass-through of changes in output-based allocations of emissions credits (see \cite{fischerfox07} and \cite{fischerfox12}) to wholesale prices, as allocation rates did not vary systematically with changes in carbon prices. We cannot directly speak to the optimal allocations of emissions credits as in \cite{burtraw_palmer_2008}, but our results suggest that there is symmetric and incomplete pass-through of the marginal value of output-based allocations and carbon prices. To our knowledge, this has not been confirmed in other studies of carbon price pass-through. There are limits to this result as well, since the changes in output-based allocations systematically disadvantaged coal plants relative to other fuel sources, and we coincident with substantial declines in natural gas prices. This implies that Alberta's coal plants were already at a substantial competitive disadvantage, and thus unable to pass through higher net carbon policy costs associated with reduced output-based allocations of emissions credits.

The Alberta wholesale power market provides an excellent quasi-laboratory for our study because it is small, isolated, and the market is settled on a real-time, energy-only basis. As such, there are few confounding factors beyond what is observable in our data. Alberta's market is winter-peaking, with peak 2021 load of just under 12,000 MW. Alberta has only three small interconnections to British Columbia, Montana, and Saskatchewan, with total capacity of less than 15\% of peak internal load.\footnote{The British Columbia and Montana tie lines have a joint operating capacity of 1500 MW for imports and 1325 MW for export, while the Saskatchewan tie line has a path rating of 150 MW.} The relative isolation of Alberta's market means that local prices and generation will be more affected by Alberta's policies than would be the case if traded power had a more important market share. The downside to this isolation is that Alberta, a major oil-producing jurisdiction, was hit with a major, oil-price-induced recession contemporaneously with some of the changes in carbon policies, and this economic downturn led to a period of low prices in the power market. Alberta's generation mix is also somewhat unique, with a significant portion of internal load served by combined heat and power (or co-generation) plants associated with industrial facilities, primarily located in the oil sands region in Northern Alberta. Despite these anomalies, the transparency and isolation of Alberta's market provide us with a unique opportunity to study carbon price pass-through to prices as well as the induced changes in market participation.

We compile over 13 years of hourly data describing market offer behaviour and a wide range of relevant characteristics. We measure and describe facility-level hourly offers of power into the market for the time period preceding the COVID-19 pandemic, although we also test this truncation of our data in later results. We combine power market data with detailed plant characteristics compiled from regulatory filings, emissions compliance data from Alberta's and Canada's air emissions reporting, weather data, commodity price information, and power market data including hourly renewable generation, imports and exports, import and export capabilities and total and forecast internal loads. Finally, we include both observed carbon policy costs and the value of output-based allocations at the facility level.

In order to isolate the impact of the changes in policies on offer behaviour, we develop a unique empirical strategy based on merit order sampling. For each of three different aggregation levels (market, plant type, and controlling entity), we build hourly merit orders of offered generation. We sample these merit orders at various support points which allows us to build a synthetic merit order with substantially-reduced dimensionality, and also to make the date comparable across time period with different total market sizes. We also include measures of marginal carbon emissions policy costs along the merit order which we sample in parallel with the offered prices. As such, we derive a sample of merit order levels and the impacts of the marginal costs of emissions policies embedded in those offers. If firms act as they are predicted to by standard economic models, passing marginal costs through into their power offers, we will see a 1:1 relationship between these costs and upward shifts in our synthetic merit orders.

@@

Finally, Alberta's market is highly concentrated, with a market structure that has evolved in ways relevant to our analysis during the relevant time period. When the market was restructured in 2000, a system of power purchase arrangements (PPAs) were used as an alternative to divestment \cite{kleit_electricity_2007}. Large coal-fired plants owned by legacy generators were covered by these PPAs which approximated the conditions for plant owners which had been present under regulation, while auctioning off the rights to offer control over these facilities to other entities. PPAs not sold in the auction were held by an entity known as the Balancing Pool. A substantial amount of Alberta's generating capacity (5066 MW) was offered into the market under these PPAs during our study period. The PPAs included a cumbersomely-worded change-of-law clause which was interpreted to imply an option for PPA owners to exit these arrangements should they become unprofitable or ``more unprofitable" due to a change in law. In response to the changes in carbon pricing policy announced in June of 2015, all PPA owners exercised rights of return \cite{bankes_termination_2016}, and offer control over all of the legacy plants (3866 MW) reverted to the Balancing Pool between January 1, 2016 and May 5, 2016 (see Table \ref{tab:PPA}). As detailed in \cite{leach_tombe_ppa}, this was caused nominally by the increase in stringency of the \textit{SGER}, but the adoption of the Climate Leadership Plan and the \textit{CCIR} meant that the expected value of the PPAs was negative for their owners \cite{leach_tombe_ppa}. The fact that these changes in offer control, and implied changes in market power shown in Figure \ref{fig:mkt_power}, happened coincidentally with changes in carbon pricing in 2016 and 2017 have the potential to confound our inference. We discuss this in our results section.

@@

Alberta's three carbon pricing regimes each affected generating facilities in two ways. First, each of three (\emph{SGER}, \emph{CCIR}, and \emph{TIER}) regimes imposes a price on emissions. Second, each regime allocates emissions credits based on output or deemed avoided emissions. The net effect of these policies creates variation within and across facilities and facility types and over time.  Consider, for example, Figure \ref{fig:sector_compliance_cost} which shows the mean value and range of annual compliance costs for larger generators in Alberta's fleet over the sample period.\footnote{Plants were only included in the sample for this figure if we had facility-specific compliance cost data. For our estimation results, we impute compliance costs for a wider set of facilities using emissions intensities from the compliance cost data.}

>>>>

%%%% AUTHOR INFORMATION %%%%
% AUTHOR INFO
\author{Andrew Leach\\
 University of Alberta}
 \begin{document}
\maketitle

@@

This paper evaluates the response of wholesale power market participants in Alberta to a series of changes in greenhouse gas (GHG) emissions policies.  Between 2015 and 2024, legislative and regulatory changes have affected carbon price levels, the rates of output-based allocations of emissions credits, and the coverage of carbon pricing across facilities of different sizes, creating variation in average and marginal carbon prices both across and within facility types and across and within portfolios held by major players in Alberta's wholesale power market. I exploit this variance in treatment to provide unique evidence of carbon price pass-through in power markets, and show that even after adjusting for costs imposed by carbon policies at the facility level, statistically significant changes in portfolio-level offer curves were induced leading to a shift in the merit order toward gas and away from coal which exceeded that which would have been predicted by carbon prices alone. These changes have accelerated the phase-out of coal-fired power in the province. My results also demonstrate that, in contrast to much of the recent literature, net marginal carbon pricing costs were not fully passed-through to offers in the wholesale market. In fact, the results of my analysis suggest that less than half of the changes in carbon policy costs was reflected in shifts in the merit order of electricity price offers in Alberta.

The traditional view of competitive firm response to carbon pricing would be an upward shift in each firm's supply function of an amount equal to the net, incremental costs imposed by the policy. For some facilities, there may be dynamic gains to participating in the market in a given hour (for example, if shutdown implies multi-hour restarts, as can be the case for coal generators), in which case the supply curve may lie below zero up to some positive quantity supplied in each hour. For these facilities, especially if they are inframarginal, we might not expect to see a carbon price change their bidding behaviour up to some threshold quantity. However, above such a threshold, we should expect to see carbon price pass-through into power offers. However, my results in the Alberta context suggest less than perfect pass-through, even for units likely to be on the margin in the market in a given hour.

The degree to which carbon prices are passed-through to wholesale electricity markets has been extensively investigated. The studies of \cite{fabra_reguant} and \cite{hintermann16_germany} are most-closely related to our analysis, and both find that carbon costs imposed in the European Union emissions trading system are almost completely passed-through to market prices. In the case of \citeauthor{fabra_reguant}, they find more-than-complete (110\%) pass-through of carbon costs in peak hours, and less-than-complete (60\%) pass-through in off-peak hours. Similarly, \citeauthor{hintermann16_germany} finds pass-through rates of 81-111\% depending on the hour of the day, and also finds that costs are fully passed-through on average. Our paper is unique in that we can measure facility-specific carbon policy costs (prices and the value of output-based allocations) precisely for long periods of time, which allows us to separately identify changes due to carbon pricing from changes due to other coincident market or climate factors. Furthermore, the changes in Alberta's climate change policies were such that the average and net marginal costs of carbon pricing policies changed across facilities for the same time periods. For example, in 2018, the marginal carbon price did not change from 2017, but there were substantial changes in output-based allocations of emissions credits such that marginal net carbon costs went up for some facilities and down for others. I use this variation to identify pass-through more precisely than is possible with daily market fluctuations in a cap-and-trade market. My results estimate much lower pass-through of carbon prices to electricity offers and thus to market prices than either \cite{fabra_reguant} or \cite{hintermann16_germany}.

Average carbon policy costs varied over time and between fuel types in Alberta through the various changes to emissions policies since 2015. This links this paper to \cite{cullen_mansur} which uses variations in fuel costs (coal vs. natural gas) to identify impacts on emissions. They find substantial impacts from small changes in fuel costs which make natural-gas-fired plants more competitive relative to coal plants. The policy changes I study in Alberta, both the changes in output-based allocations and the carbon price increases, have had a similar effect on the coal-to-gas cost ratio as a reduction in natural gas prices. Their results also inform my results in the sense that the same gas price reductions seen in their study occurred in Alberta during our sample period, and we adjust for these impacts in our study design. I show that the degree to which carbon prices were passed through was most consistent with the theory for more flexible, combined-cycle natural gas facilities, while coal plants passed through a much lower share of increased emissions costs.

The Alberta electricity market is highly-concentrated, which links our paper to the literature following from the seminal paper of \cite{buchanan1969external} on externalities and related policies in the presence of market power, as well as to myriad studies of market power in electricity, e.g. \cite{borenstein_bushnell_wolak_2002}. For example, \cite{fowlie_reguant_ryan_2016} study the interaction of market structure in the cement industry and greenhouse gas mitigation policies. \cite{miller_cement}, also looking at the cement industry, shows that fuel cost changes were more-than-fully reflected in downstream pricing. And, as in \cite{holland_emissions_2012}, incomplete regulation or regulations which provide competing incentives will distort price-pass through. The latter also applies in the present study since there are changes in market structure coincident with the policy changes we study for which we must account.

Another unique attribute of this study is that the policy variation over time and across facilities in Alberta also allows us to directly identify the pass-through of changes in output-based allocations of emissions credits (see \cite{fischerfox07} and \cite{fischerfox12}) to wholesale prices, as allocation rates did not vary systematically with changes in carbon prices. The analysis cannot directly speak to the optimal allocations of emissions credits as in \cite{burtraw_palmer_2008}, but my results suggest that there is symmetric and incomplete pass-through of the marginal value of output-based allocations and carbon prices. To my knowledge, this has not been confirmed in other studies of carbon price pass-through. There are limits to this result as well, since the changes in output-based allocations systematically disadvantaged coal plants relative to other fuel sources, and we coincident with substantial declines in natural gas prices. This implies that Alberta's coal plants were already at a substantial competitive disadvantage, and thus unable to pass through higher net carbon policy costs associated with reduced output-based allocations of emissions credits.

The Alberta wholesale power market provides an excellent quasi-laboratory for this study because it is small, isolated, and the market is settled on a real-time, energy-only basis. As such, there are few confounding factors beyond what is observable in publicly-available data. Alberta's market is winter-peaking, with peak load of just over 12,000 MW. Alberta has only three small interconnections to British Columbia, Montana, and Saskatchewan, with total capacity of less than 15\% of peak internal load.\footnote{The British Columbia and Montana tie lines have a joint operating capacity of 1500 MW for imports and 1325 MW for export, while the Saskatchewan tie line has a path rating of 150 MW.} The relative isolation of Alberta's market means that local prices and generation will be more affected by Alberta's policies than would be the case if traded power had a more important market share. The downside to this isolation is that Alberta, a major oil-producing jurisdiction, was hit with a major, oil-price-induced recession contemporaneously with some of the changes in carbon policies, and this economic downturn led to a period of low prices in the power market. Alberta's generation mix is also somewhat unique, with a significant portion of internal load served by combined heat and power (or co-generation) plants associated with industrial facilities, primarily located in the oil sands region in Northern Alberta. Despite these anomalies, the transparency and isolation of Alberta's market provide us with a unique opportunity to study carbon price pass-through to prices as well as the induced changes in market participation.

The analysis in this paper is based on over 15 years of hourly data describing market offer behaviour and a wide range of relevant characteristics. We combine power market data with detailed plant characteristics compiled from regulatory filings, emissions compliance data from Alberta's and Canada's air emissions reporting, weather data, commodity price information, and power market data including hourly renewable generation, imports and exports, import and export capabilities, and total and forecast internal loads. Finally, we include both observed carbon policy costs and the value of output-based allocations at the facility level.

In order to isolate the impact of the changes in policies on offer behaviour, we develop a unique empirical strategy based on merit order sampling. For each of three different aggregation levels (market, plant type, and controlling entity), we build hourly merit orders of offered generation. We sample these merit orders at various support points which allows us to build a synthetic merit order with substantially-reduced dimensionality, and also to make the data comparable across time periods with different total market sizes. We also include measures of marginal carbon emissions policy costs along the merit order which we sample in parallel with the offered prices. As such, we derive a sample of merit order levels and the impacts of the marginal costs of emissions policies embedded in those offers. If firms act as they are predicted to by standard economic models, passing marginal costs through into their power offers, we will see a 1:1 relationship between these costs and upward shifts in our synthetic merit orders.

@@

Finally, Alberta's market is highly-concentrated, with a market structure that has evolved in ways relevant to our analysis during our sample period. When the market was restructured in 2000, a system of power purchase arrangements (PPAs) were used as an alternative to divestment \cite{kleit_electricity_2007}. Large coal-fired plants owned by legacy generators were covered by these PPAs which approximated the conditions for plant owners which had been present under regulation, while auctioning off the rights to offer control over these facilities to other entities. PPAs not sold in the auction were held by an entity known as the Balancing Pool. A substantial amount of Alberta's generating capacity (5066 MW) was offered into the market under these PPAs during our study period. The PPAs included a cumbersomely-worded change-of-law clause which was interpreted to imply an option for PPA owners to exit these arrangements should they become unprofitable or ``more unprofitable" due to a change in law. In response to the changes in carbon pricing policy announced in June of 2015, all private PPA owners exercised rights of return \cite{bankes_termination_2016}, and offer control over all of the legacy plants (3866 MW) reverted to the Balancing Pool between January 1, 2016 and May 5, 2016 (see Table \ref{tab:PPA}). As detailed in \cite{leach_tombe_ppa}, this was caused nominally by the increase in stringency of the \textit{SGER}, but the adoption of the Climate Leadership Plan and the \textit{CCIR} meant that the expected value of the PPAs was negative for their owners \cite{leach_tombe_ppa}. The fact that these changes in offer control, and implied changes in market power shown in Figure \ref{fig:mkt_power}, happened coincidentally with changes in carbon pricing in 2016 and 2017 have the potential to confound our inference. We discuss this in our results section.

@@

Alberta's three carbon pricing regimes each affected generating facilities through two channels. First, each of the three (\emph{SGER}, \emph{CCIR}, and \emph{TIER}) regimes imposes a price on emissions. Second, each regime allocated emissions credits based on output or deemed avoided emissions. The net effects of these policies create variation within and across facilities and facility types and over time.  Consider, for example, Figure \ref{fig:sector_compliance_cost} which shows the mean value and range of annual compliance costs for larger generators in Alberta's fleet over the sample period.\footnote{Plants were only included in the sample for this figure if we had facility-specific compliance cost data. For our estimation results, we impute compliance costs for a wider set of facilities using emissions intensities from the compliance cost data.}
